---
title: "Census Data Access"
author: "Sheridamae Bondoc Gudez, M.S."
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=F}
knitr::opts_chunk$set(echo = FALSE, message=T, warning=T)
rm(list=ls())
#Read in installed package list
libraries <- c("tidyverse", "tidycensus", "sf", "tigris", "sp", "knitr")
for (i in libraries){
  tryCatch(library(i, character.only = TRUE),
           print(paste(i, "installed")),
           error = function(e) {install.packages(i);
             library(i, character.only = TRUE)}
  )
  print(paste(i, "loaded"))
}
options(tigris_class = "sf")
options(scipen = 999)
```

# Census API Key

To access the US Census Database, you first must have a unique API key which you can request for free here: <https://api.census.gov/data/key_signup.html>

An API key will then be emailed to you in a few hours to a few days, depending on how much traffic the site is getting.

```{r api}
census_api_key("api_key") #insert api key
```

# Getting ACS Codebook

```{r acs_codes, echo= T}
acs_variables <- load_variables(2020, "acs5", cache = TRUE) #pull codebook for the 5-year ACS database. Change year to desired timeframe (i.e., 2020 5-year ACS ranges from 2016 to 2020)
```

# Accessing Census Data and Creating Desired Variables

## Custom Pull

```{r}
acs <- get_acs(geography = "tract", #unit of analysis (i.e, tract, county, block)
                  year = 2020, #desired year
                  variables = "variable_name" = "variable_code", #custom variable name and the variable code from the codebook
                  state = "state_abv", #desired state
                  county = "county_name", #desired county
                  survey = "acs5", #desired census database, here we us 5-year ACS
                  geometry = TRUE) %>% #allow for spatial data to be included
                  select(-(moe)) %>%
                  spread(key = variable, value = estimate)
```

## Common Community Research Variables

```{r full_census}
acs <- get_acs(geography = "tract", 
                  year = 2020,
                  variables = c(tpop= "B01003_001", #total population
                                
                                white= "B03002_003", #total non-hispanic white pop.
                                black= "B03002_004", #total non-hispanic black pop.
                                asian= "B03002_006", #total non-hispanic asian pop.
                                hisp= "B03002_012", #total hispanic pop.
                                ntvam= "B03002_005", #total non-hispanic native american pop.
                                pacif= "B03002_007", #total non-hispanic pacific islander pop.
                                oth1= "B03002_009", #total of two or more races pop.
                                oth2= "B03002_008", #total of some other race
                                
                                fborn= "B99051_005", #total foreign born
                                
                                singmm= "B09002_015", #total single mothers (no spouse present and with children under 18)
                                hsnocomp= "B15003_002", #total who did not receive a high school degree
                                poverty= "B17010_002", #total under the poverty line
                                medinc= "B19013_001", #median income
                                unply= "B23025_007", #total unemployed that are aged 16 to 64
                                snap= "B22010_002", #Total households on Food Stamps/SNAP
                                assist= "B19057_002", #Total households on public assistance
                                
                                renter= "B25003_003", #total renter occupied households
                                rsyr= "B25039_001", #average move-in year of household
                                
                                m1517= "B01001_006", #males aged 15-17
                                m1819= "B01001_007", #males aged 18-19
                                m20= "B01001_008", #males aged 20
                                m21= "B01001_009", #males aged 21
                                m2224= "B01001_010", #males aged 22-24
                                
                                marr= "B09021_003" #total married 
                                ), 
                  state = "PA",
                  county = "Centre County",
                  survey = "acs5",
                  geometry = TRUE) %>%
  select(-(moe)) %>%
  spread(key = variable, value = estimate) %>%
  mutate(area_km2 = st_area(geometry) / 1e6,  # Area in square kilometers
         ppdense = as.numeric(tpop / area_km2),   # Population density (people per km²)
         tpop_log = log(tpop),
         
         oth= ntvam+pacif+oth1+oth2, #other pop.
         pwht= (white/tpop), #percent white
         pblk=(black/tpop), #percent black
         pasn=(asian/tpop), #percent asian
         phisp= (hisp/tpop), #percent hispanic
         poth= (oth/tpop), #percent other
         tpopr= white+black+asian+hisp+oth, #total races pop.
         
         pfbrn= (fborn/tpop), #percent foreign born
         
         pass= ((snap+assist)/tpop), #Percent households on public assistance
         phsno= hsnocomp/tpop, #Percent without a high school diploma
         medinc_log= log(medinc), #median income (logged)
         puply= (unply/tpop), #percent unemployed
         ppov= (poverty/tpop), #percent under the poverty line
         psinm= (singmm/tpop), #percent single mother
         
         prent= (renter/tpop), #percent renters
         avgrs= abs((2017-rsyr)), #average tract residency length end year of ACS - average year of residency)
         
         p1524= ((m1517+m1819+m20+m21+m2224)/tpop),
         pmarr= (marr/tpop) #percent married couples
         ) %>% 
                  filter(tpop != 0) #removing units with no populations (i.e airports, parks)
```

# Clip Data for a Desired City

```{r}
pl <- places(state = "PA", cb = TRUE, year=2020) # Bring in 2020 city boundary data. Change year if using a different version of the ACS data
city <- filter(pl, grepl("State College", NAME)) #filter specified city
city_map <- st_intersection(acs, city) #Clip data using city boundary
```

# Saving Non-Spatial Dataframe and a Shapefile

## Pre-Processing

```{r}
#setting up dataframe for further cleaning
data <- as.data.frame(maps) #remove spatial element and convert to data frame
data <- data %>% #remove geometry and other spatial variables
  select(GEOID, ppdense, tpop, 
                    pwht, pblk, pasn, phisp, poth, 
                    white, black, asian, hisp, oth, 
                    medinc_log, ppov, puply, phsno, pass, psinm, p1524, 
                    pvac, phmwn, avgrs, homeown) #change variables to desired ones

#setting up shapefile for mapping
maps <- maps %>%
  select(c(GEOID, geometry)) #remove all variables except for geometry and GEOIDs
```

## Saving

```{r}
write.csv(data, "city_df.csv") #save as .csv file 
write_sf(maps, "city.map.shp", delete_layer = T) #save shapefile and OVERWRITE previous one if detected using delete_layer=T
```

# References

Pebesma, E., 2018.
Simple Features for R: Standardized Support for Spatial Vector Data.
The R Journal 10 (1), 439-446, <https://doi.org/10.32614/RJ-2018-009>

Pebesma, E., & Bivand, R.
(2023).
Spatial Data Science: With Applications in R.
Chapman and Hall/CRC.
<https://doi.org/10.1201/9780429459016>

Walker K, Herman M (2025). 
*tidycensus: Load US Census Boundary and Attribute Data as 'tidyverse' and 'sf'-Ready Data Frames*.
R package version 1.7.1, <https://walker-data.com/tidycensus/>.
